Computational Methods for Principle Component Analysis Applied to Genotype Data

### 2018-04-30 doc/paper.tex

\usepackage{epsfig} \usepackage[margin=1.5in]{geometry}
\usepackage[semicolon,authoryear]{natbib} \bibliographystyle{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cleveref}
\usepackage{verbatim}
\usepackage{graphicx}

@@

The standard way of predicting PC scores is to find the PC loadings of the reference data and project the data points of the study samples onto them.
We call this method simple projection.
However, when the number of features are greatly larger than the size of the reference sample,
the predicted PC scores by this method are known to be biased toward null.
One way of addressing this shrinkage bias is presented by \cite{wang}.
Their solution is to combine one study individual with the reference set and
find the PC scores of this augmented data set by singular value decomposition (SVD).
The PC scores of the study individuals are then retrieved from the singular vector matrices and mapped to the reference PC space by a Procrustes transformation.
We call this method ``augment, decompose, and transform'' (ADT).
This method has been shown to be effective in eliminating the shrinking bias of
study PC scores.
However since ADT needs to run SVD for each study samples, ADT is computationally expensive.

To address the limitations of simple projection approach and ADT, we propose and evaluate two alternative PC score prediction methods.
The first approach removes the bias in the simple projection method by estimating the asymptotic bias factor, which is calculated based on Random Matrix Theory \cite{dey}.
The second approach improves the computation efficiency of ADT by using online SVD algorithm. 
We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment-decompose-transform'' (OADT).

@@

In this paper, we compare the properties of
simple projection, augment-decompose-transform, adjusted projection, and online augment-decompose-transform
in terms of computation time and accuracy.
We also compare using standard SVD and randomized SVD for the reference set
and examine the difference in accuracy and computation time.
We found that HDPCA projection and OADT have both achieved the accuracy of ADT
and the computational efficiency of simple projection.
In addition, randomized SVD gives results similar to standard SVD and only increase the computational time for the reference data.

@@

\subsubsection{Augment, Decompose, and Transform (ADT)}

One way to solve the shrinking bias of SP is the ADT approach developed by \cite{wang}.
In this method, each study individual's data point is appended to the reference
data matrix.
Then SVD is applied to the augmented matrix,
and PC scores in this augmented sample are found for both the reference
individuals and the study individual.
The the PC scores of the reference individuals in this augmented sample is
compared with the PC scores of the reference individuals in the reference
sample,
which is calculated in the beginning,
and a linear (Procrustes) map is constructed to transform the latter into the former
with the smallest error.
This transformation is then applied to the PC scores of the study individual in
the augmented sample.
The resulted PC scores are used as an estimate for the study individual's
population PC scores.
Specifically, let
\[
  V =
  \begin{bmatrix}
    V_{\text{ref}} & W
  \end{bmatrix},
  \quad
  \tilde{V} =
  \begin{bmatrix}
    \tilde{V}_{\text{ref}} & \tilde{W}_\text{ref} \\
    \tilde{V}_\text{stu} & \tilde{W}_\text{stu}
  \end{bmatrix},
\]
where the dimensions of $V_\text{ref}$, $\tilde{V}_\text{ref}$, and $\tilde{V}_\text{stu}$ are $n \times k_1$, $n \times k_2$, and $1 \times k_2$, respectively, with $ 1 \leq k_1 \leq k_2 \leq n+1$. 
Next, we use Procrustes analysis to find
\[
  A, b = \argmin{A, b} d(V_\text{ref}, \tilde{V}_{\text{ref}}A+b)
\]
for some metric $d$.
Finally, we get the Procrustes-adjusted top $k_1$ PC scores for the study individual $V_\text{pro} = \tilde{V}_\text{stu} A + b$.
The algorithm is summarized as follows:
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$ (columns of $D$ and $V$ are sorted by PC variance with the first column corresponding to the greatest PC variance).
  ($\bO[n^3]$.)
\item Append $X^T y$ and its transpose to the last column and row of $X^T X$, respectively.
  The entry of the bottom-right corner is $y^T y$.
  This is exactly the same as computing $\tilde{X}^T \tilde{X}$,
  where $\tilde{X} = (X, y)$ is the reference group with the study individual added.
  ($\bO[np]$.)
\item Apply eigendecomposition to $\tilde{X}^T \tilde{X}$ to get $\tilde{X}^T \tilde{X} = \tilde{V} \tilde{D}^2 \tilde{V}^T$.
  ($\bO[n^3]$.)
\item Project the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual (the last row of $V$),
  where $k_1 < k_2 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 3 for the next study individual.
\end{enumerate}

@@

Although accurate, ADT is the slowest among the other three computational methods.
Its computational complexity increases cubically as the reference sample size and linearly as the study sampl sizes.
The ADT method is has been implemented in the TRACE software by \cite{wang}.

@@

\subsubsection{Online ADT (OADT)}

A remedy for ADT is to reduce the computational complexity without much sacrifice of the accuracy.
This is achieved by the online ADT (OADT) method.
Notice that in OADT, only the top $k_2$ PC's are used.
OADT uses the online SVD method, which calculates a close approximation of the first $k_3$ columns of $\tilde{V}$ for some $k_2 \leq k_3 \ll n$ and keeps everything else the same as in ADT, including Procrustes analysis. 
The algorithm is
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $U_3 = X V_3 D_3^{-1}$,
  where $V_3$ is the first $k_3$ columns of $V$,
  and $D_3$ is the diagonal matrix of the first $k_3$ diagonal entries of $D$.
  ($\bO[k_3 n p]$.)
\item Calculate 
  \[
    L = U_3^T y \quad \text{and} \quad K = y^T H,
  \]
  where $H$ is the normalized  $y - U_3L$
  ($\bO[k_3 p]$.)
\item Calculate $Q^T Q$, where
  \[
    Q = 
    \begin{bmatrix}
      D_3 & L \\
      0 & K
    \end{bmatrix}.
  \]
  ($\bO[k_3^3]$.)
\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V}_3 \ddot{D}^2_3 \ddot{V}^T_3$.
  ($\bO[k_3^3]$.)
\item Calculate
  \[
    \tilde{V}_3 =
    \begin{bmatrix}
      V_3 & 0 \\
      0 & 1
    \end{bmatrix}
    \ddot{V}_3.
  \]
  ($\bO[nk_3^2]$.)
\item Project the first $k_2$ columns of $\tilde{V}_3$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual,
  where $k_1 < k_2  < k_3 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 4 for the next study individual.
\end{enumerate}

@@

The computational complexity of OADT for the study individuals increases linearly with respect to the reference sample size, which makes the computation time of OADT close to those of SP.
The closeness between the results given by OADT and ADT will be empirically shown later.

@@

In addition to estimating the PC scores of the study individuals,
we have another algorithm that aims at improving the decomposition of the
training data.
For SP, AP, ADT, and OADT, we need to find the covariance matrix of the training
data, which has dimension $n \times n$, and conduct eigendecomosition on it.
However, when $n$ is exceedingly large,
computation and memory cost of PCA on the reference samples can be enormous.
A randomized SVD (RSVD) algorithm has been developed to handle this issue.
When the reference sample size is small, RSVD can be slower than standard SVD,
but it is scalable to large sample size.
For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.

@@

For comparing the accuracy, we use two different measures.
First, for a given cell of the grid, we compute the mean of the PC scores of the 
reference individuals located at this cell.  
This is the estimated coordinate of the cell.
Next, we find the Euclidean distance between the estimated PC scores of the
study individuals and the estimated coordinates of the cells they are located
at.
These are the geographic estimation errors for the study individuals.
The average of these errors are used to measure the accuracy of the
computational methods (that is, the square root of the sum of squares divided by the
study size). 
Second, since ADT is a model free approach and believed to be the most reliable, we compare the average
Euclidean distance between the PC score estimates calculated by another method
to those calculated by ADT. 
The comparison of the runtimes and accuracy are also conducted by using both the
standard SVD and randomized SVD on the training data matrix.

@@

For the real-life datasets (\Cref{tbl:ukb} and \Cref{fig:ukb}),
we see that ADT, which is the most reliable method,
is the slowest in study runtimes.
By using the online SVD algorithm,
OADT achieved the highest accuracy (excluding ADT, which is the golden standard) and the fastest runtime.
SP is fast, also,
but its accuracy is the lowest.
By using the HDPCA package,
AP reduces the error of AP
and achieved the second highest in accuracy.
The randomized version of SP is as accurate as the standard SP,
which shows that the randomized SVD algorithm gives reliable results,
and that using linear regression to predict the uncalculated reference eigenvalues is satisfactory.
It also reduces the runtime almost by half,
due to not calculating all the eigenvalues.
Similar reduction in study runtime is observed for the randomized version of AP,
but in this case the accuracy is also decreased,
thought it is still higher than that of SP and RSP.
Finally, the randomized OADT is disappointing,
as its error is almost as high as SP and RSP.

The pattern for accuracy is very similar when we look at the simulated data.
(\Cref{tbl:nChg-runtimes-ref}, \Cref{tbl:nChg-runtimes-study},
\Cref{tbl:nChg-accuracy-gold}, \Cref{tbl:nChg-accuracy-ctr},
\Cref{tbl:mChg-runtimes-study}).
A noticeable difference in this case is the accuracy of ROADT.
Here ROADT's accuracy is almost identical to OADT's,
as measured by both the golden standard error and the geographic center err.
However, as the reference size increases,
the accuracies of different methods become closer and closer (\Cref{fig:nChg}).
As for the speed,
AWD's study runtime increases so fast with the reference size
that the study runtimes for the other methods seem to be constant in comparison (\Cref{fig:nChg}).
On the other hand,
the study runtimes appear to increase linearly with respect to the study size (\Cref{fig:mChg}).
This information can be used to predict the runtime for larger study samples.

@@

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|}
    \hline
    Method & Reference Complexity & Study Complexity \\ 
    \hline
    ADT & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
    \hline
    OADT & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
    \hline
    ROADT & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
    \hline
    SP & $\bO[n^2p]$ & $\bO[mkp]$ \\
    \hline
    RSP & $\bO[k^2p]$ & $\bO[mkp]$ \\
    \hline
    AP & $\bO[n^2p]$ &  $\bO[mkp]$ \\
    \hline
    RAP & $\bO[k^2p]$ &  $\bO[mkp]$ \\
    \hline
  \end{tabular}
  \caption{
    Comparison of computational complexity for different methods.
    Here $p$ is the number of SNPs,
    $n$ is the number of individuals in the reference group,
    $m$ is the number of individuals in the  study group,
    and $k$ is the number of PCs calculated by the corresponding method.
  }
  \label{tbl:cplx}
\end{table}

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|l|l|}
    \hline
    Method & Reference Runtimes (n=?) & Study Runtimes (m=?)& Golden Standard Error \\ 
    \hline
    ADT & 411 & 29813 & 0.00 \\
    \hline
    OADT & 2518 & 103 & 0.40 \\
    \hline
    ROADT & 2518 & 103 & 1.93 \\
    \hline
    SP & 411 & 425 & 1.95 \\
    \hline
    RSP & 2504 & 241 & 1.95 \\
    \hline
    AP & 411 & 482 & 1.12 \\
    \hline
    RAP & 2505 & 250 & 1.83 \\ 
    \hline
  \end{tabular}
  \caption{
    Comparison of runtimes and accuracy for different methods applied to the real-life datasets.
    The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals.
    The study sample contain 1000 individuals from the UKBioBank dataset.
    There are about 125,000 loci shared by the reference and study data.
    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADT.
  }
  \label{tbl:ukb}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        &  68  & 1263 & 2871 & 5720 & 9737 \\
    OADT       &  72  & 18   & 19   & 22   & 24   \\
    ROADT      & 809  & 18   & 17   & 20   & 21   \\
    SP         &  68  & 26   & 42   & 67   & 96   \\
    RSP        & 805  & 23   & 31   & 39   & 46   \\
    AP         &  68  & 69   & 107  & 91   & 119  \\
    RAP        & 805  & 11   & 24   & 29   & 9    \\
    \hline
  \end{tabular}
  \caption{
    Reference runtimes for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-runtimes-ref}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        & 391  &  130 &  222 &  339 &  486 \\
    OADT       & 15   &  136 &  230 &  350 &  499 \\
    ROADT      & 15   & 1211 & 1633 & 2023 & 2422 \\
    SP         & 16   &  130 &  222 &  339 &  486 \\
    RSP        & 15   & 1204 & 1624 & 2012 & 2409 \\
    AP         & 34   &  130 &  222 &  339 &  486 \\
    RAP        & 18   & 1204 & 1624 & 2012 & 2409 \\
    \hline
  \end{tabular}
  \caption{
    Study runtimes for simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-runtimes-study}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Study Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        & 0.00  & 0.00 & 0.00 & 0.00 & 0.00 \\
    OADT       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
    ROADT      & 0.23  & 0.15 & 0.12 & 0.08 & 0.08 \\
    SP         & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
    RSP        & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
    AP         & 0.46  & 0.37 & 0.31 & 0.26 & 0.22 \\
    RAP        & 3.02  & 2.93 & 2.54 & 2.15 & 1.91 \\
    \hline
  \end{tabular}
  \caption{
    Golden standard error for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-accuracy-gold}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        & 4.53  & 4.13 & 3.81 & 3.59 & 3.23 \\
    OADT       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
    ROADT      & 4.44  & 4.06 & 3.76 & 3.55 & 3.20 \\
    SP         & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
    RSP        & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
    AP         & 4.59  & 4.10 & 3.76 & 3.57 & 3.24 \\
    RAP        & 4.23  & 3.87 & 3.55 & 3.38 & 3.07 \\
    \hline
  \end{tabular}
  \caption{
    Geographic center error for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-accuracy-ctr}
\end{table}

>>>>

\usepackage{epsfig} \usepackage[margin=1.5in]{geometry}
\usepackage[semicolon,authoryear]{natbib} \bibliographystyle{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cleveref}
\usepackage{verbatim}
\usepackage{float}

@@

The standard way of predicting PC scores is to find the PC loadings of the reference data and project the data points of the study samples onto them.
We call this method simple projection (SP).
However, when the number of features are greatly larger than the size of the reference sample,
the predicted PC scores by this method are known to be biased toward null.
One way of addressing this shrinkage bias is presented by \cite{wang}.
Their solution is to combine one study individual with the reference set and
find the PC scores of this augmented data set by singular value decomposition (SVD).
The PC scores of the study individuals are then retrieved from the singular vector matrices and mapped to the reference PC space by a Procrustes transformation.
We call this method ``augment, decompose, and Procrustes transform'' (ADP).
This method has been shown to be effective in eliminating the shrinking bias of
study PC scores.
However since ADP needs to run SVD for each study samples, ADP is computationally expensive.

To address the limitations of SP and ADP,
we propose and evaluate two alternative PC score prediction methods.
The first approach removes the bias in SP by estimating the asymptotic bias factor,
which is calculated based on Random Matrix Theory \cite{dey}.
The second approach improves the computation efficiency of ADP by using online SVD algorithm. 
We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment, decompose, and and Procrustes transform'' (OADP).

@@

In this paper, we compare the properties of
SP, ADP, AP, and OADP in terms of computation time and accuracy.
We also compare using standard SVD and randomized SVD for the reference set
and examine the difference in accuracy and computation time.
We found that AP and OADP have both achieved the accuracy of ADP
and the computational efficiency of SP.
In addition, randomized SVD gives results similar to standard SVD and only increase the computational time for the reference data.

@@

\subsubsection{Augment, Decompose, and Transform (ADP)}

One way to solve the shrinking bias of SP is the ADP approach developed by \cite{wang}.
In this method, each study individual's data point is appended to the reference
data matrix.
Then SVD is applied to the augmented matrix,
and PC scores in this augmented sample are found for both the reference
individuals and the study individual.
The the PC scores of the reference individuals in this augmented sample is
compared with the PC scores of the reference individuals in the reference
sample,
which is calculated in the beginning,
and a linear (Procrustes) map is constructed to transform the latter into the former
with the smallest error.
This transformation is then applied to the PC scores of the study individual in
the augmented sample.
The resulted PC scores are used as an estimate for the study individual's
population PC scores.
Specifically, let
\[
  V =
  \begin{bmatrix}
    V_{\text{ref}} & W
  \end{bmatrix},
  \quad
  \tilde{V} =
  \begin{bmatrix}
    \tilde{V}_{\text{ref}} & \tilde{W}_\text{ref} \\
    \tilde{V}_\text{stu} & \tilde{W}_\text{stu}
  \end{bmatrix},
\]
where the dimensions of $V_\text{ref}$, $\tilde{V}_\text{ref}$, and $\tilde{V}_\text{stu}$ are $n \times k_1$, $n \times k_2$, and $1 \times k_2$, respectively, with $ 1 \leq k_1 \leq k_2 \leq n+1$. 
Next, we use Procrustes analysis to find
\[
  A, b = \argmin{A, b} d(V_\text{ref}, \tilde{V}_{\text{ref}}A+b)
\]
for some metric $d$.
Finally, we get the Procrustes-adjusted top $k_1$ PC scores for the study individual $V_\text{pro} = \tilde{V}_\text{stu} A + b$.
The algorithm is summarized as follows:
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$ (columns of $D$ and $V$ are sorted by PC variance with the first column corresponding to the greatest PC variance).
  ($\bO[n^3]$.)
\item Append $X^T y$ and its transpose to the last column and row of $X^T X$, respectively.
  The entry of the bottom-right corner is $y^T y$.
  This is exactly the same as computing $\tilde{X}^T \tilde{X}$,
  where $\tilde{X} = (X, y)$ is the reference group with the study individual added.
  ($\bO[np]$.)
\item Apply eigendecomposition to $\tilde{X}^T \tilde{X}$ to get $\tilde{X}^T \tilde{X} = \tilde{V} \tilde{D}^2 \tilde{V}^T$.
  ($\bO[n^3]$.)
\item Project the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual (the last row of $V$),
  where $k_1 < k_2 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 3 for the next study individual.
\end{enumerate}

@@

Although accurate, ADP is the slowest among the other three computational methods.
Its computational complexity increases cubically as the reference sample size and linearly as the study sampl sizes.
The ADP method is has been implemented in the TRACE software by \cite{wang}.

@@

\subsubsection{Online ADP (OADP)}

A remedy for ADP is to reduce the computational complexity without much sacrifice of the accuracy.
This is achieved by the online ADP (OADP) method.
Notice that in OADP, only the top $k_2$ PC's are used.
OADP uses the online SVD method, which calculates a close approximation of the first $k_3$ columns of $\tilde{V}$ for some $k_2 \leq k_3 \ll n$ and keeps everything else the same as in ADP, including Procrustes analysis. 
The algorithm is
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $U_3 = X V_3 D_3^{-1}$,
  where $V_3$ is the first $k_3$ columns of $V$,
  and $D_3$ is the diagonal matrix of the first $k_3$ diagonal entries of $D$.
  ($\bO[k_3 n p]$.)
\item Calculate 
  \[
    L = U_3^T y \quad \text{and} \quad K = y^T H,
  \]
  where $H$ is the normalized  $y - U_3L$
  ($\bO[k_3 p]$.)
\item Calculate $Q^T Q$, where
  \[
    Q = 
    \begin{bmatrix}
      D_3 & L \\
      0 & K
    \end{bmatrix}.
  \]
  ($\bO[k_3^3]$.)
\item Apply eigendecomosition to $Q^T Q$ to get $Q^T Q = \ddot{V}_3 \ddot{D}^2_3 \ddot{V}^T_3$.
  ($\bO[k_3^3]$.)
\item Calculate
  \[
    \tilde{V}_3 =
    \begin{bmatrix}
      V_3 & 0 \\
      0 & 1
    \end{bmatrix}
    \ddot{V}_3.
  \]
  ($\bO[nk_3^2]$.)
\item Project the first $k_2$ columns of $\tilde{V}_3$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual,
  where $k_1 < k_2  < k_3 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 4 for the next study individual.
\end{enumerate}

@@

The computational complexity of OADP for the study individuals increases linearly with respect to the reference sample size, which makes the computation time of OADP close to those of SP.
The closeness between the results given by OADP and ADP will be empirically shown later.

@@

In addition to estimating the PC scores of the study individuals,
we have another algorithm that aims at improving the decomposition of the
training data.
For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
data, which has dimension $n \times n$, and conduct eigendecomosition on it.
However, when $n$ is exceedingly large,
computation and memory cost of PCA on the reference samples can be enormous.
A randomized SVD (RSVD) algorithm has been developed to handle this issue.
When the reference sample size is small, RSVD can be slower than standard SVD,
but it is scalable to large sample size.
For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.

@@

For comparing the accuracy, we use two different measures.
First, for a given cell of the grid, we compute the mean of the PC scores of the 
reference individuals located at this cell.  
This is the estimated coordinate of the cell.
Next, we find the Euclidean distance between the estimated PC scores of the
study individuals and the estimated coordinates of the cells they are located
at.
These are the geographic estimation errors for the study individuals.
The average of these errors are used to measure the accuracy of the
computational methods (that is, the square root of the sum of squares divided by the
study size). 
Second, since ADP is a model free approach and believed to be the most reliable, we compare the average
Euclidean distance between the PC score estimates calculated by another method
to those calculated by ADP. 
The comparison of the runtimes and accuracy are also conducted by using both the
standard SVD and randomized SVD on the training data matrix.

@@

For the real-life datasets (\Cref{tbl:ukb} and \Cref{fig:ukb}),
we see that ADP, which is the most reliable method,
is the slowest in study runtimes.
By using the online SVD algorithm,
OADP achieved the highest accuracy (excluding ADP, which is the golden standard) and the fastest runtime.
SP is fast, also,
but its accuracy is the lowest.
By using the HDPCA package,
AP reduces the error of AP
and achieved the second highest in accuracy.
The randomized version of SP is as accurate as the standard SP,
which shows that the randomized SVD algorithm gives reliable results,
and that using linear regression to predict the uncalculated reference eigenvalues is satisfactory.
It also reduces the runtime almost by half,
due to not calculating all the eigenvalues.
Similar reduction in study runtime is observed for the randomized version of AP,
but in this case the accuracy is also decreased,
thought it is still higher than that of SP and RSP.
Finally, the randomized OADP is disappointing,
as its error is almost as high as SP and RSP.

The pattern for accuracy is very similar when we look at the simulated data.
(\Cref{tbl:nChg-runtimes-ref}, \Cref{tbl:nChg-runtimes-study},
\Cref{tbl:nChg-accuracy-gold}, \Cref{tbl:nChg-accuracy-ctr},
\Cref{tbl:mChg-runtimes-study}).
A noticeable difference in this case is the accuracy of ROADP.
Here ROADP's accuracy is almost identical to OADP's,
as measured by both the golden standard error and the geographic center err.
However, as the reference size increases,
the accuracies of different methods become closer and closer (\Cref{fig:nChg}).
As for the speed,
AWD's study runtime increases so fast with the reference size
that the study runtimes for the other methods seem to be constant in comparison (\Cref{fig:nChg}).
On the other hand,
the study runtimes appear to increase linearly with respect to the study size (\Cref{fig:mChg}).
This information can be used to predict the runtime for larger study samples.

@@

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|}
    \hline
    Method & Reference Complexity & Study Complexity \\ 
    \hline
    ADP & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
    \hline
    OADP & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
    \hline
    ROADP & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
    \hline
    SP & $\bO[n^2p]$ & $\bO[mkp]$ \\
    \hline
    RSP & $\bO[k^2p]$ & $\bO[mkp]$ \\
    \hline
    AP & $\bO[n^2p]$ &  $\bO[mkp]$ \\
    \hline
    RAP & $\bO[k^2p]$ &  $\bO[mkp]$ \\
    \hline
  \end{tabular}
  \caption{
    Comparison of computational complexity for different methods.
    Here $p$ is the number of SNPs,
    $n$ is the number of individuals in the reference group,
    $m$ is the number of individuals in the  study group,
    and $k$ is the number of PCs calculated by the corresponding method.
  }
  \label{tbl:cplx}
\end{table}

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|l|l|}
    \hline
    Method & Reference Runtimes (n=?) & Study Runtimes (m=?)& Golden Standard Error \\ 
    \hline
    ADP & 411 & 29813 & 0.00 \\
    \hline
    OADP & 2518 & 103 & 0.40 \\
    \hline
    ROADP & 2518 & 103 & 1.93 \\
    \hline
    SP & 411 & 425 & 1.95 \\
    \hline
    RSP & 2504 & 241 & 1.95 \\
    \hline
    AP & 411 & 482 & 1.12 \\
    \hline
    RAP & 2505 & 250 & 1.83 \\ 
    \hline
  \end{tabular}
  \caption{
    Comparison of runtimes and accuracy for different methods applied to the real-life datasets.
    The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals.
    The study sample contain 1000 individuals from the UKBioBank dataset.
    There are about 125,000 loci shared by the reference and study data.
    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADP.
  }
  \label{tbl:ukb}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADP        &  68  & 1263 & 2871 & 5720 & 9737 \\
    OADP       &  72  & 18   & 19   & 22   & 24   \\
    ROADP      & 809  & 18   & 17   & 20   & 21   \\
    SP         &  68  & 26   & 42   & 67   & 96   \\
    RSP        & 805  & 23   & 31   & 39   & 46   \\
    AP         &  68  & 69   & 107  & 91   & 119  \\
    RAP        & 805  & 11   & 24   & 29   & 9    \\
    \hline
  \end{tabular}
  \caption{
    Reference runtimes for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-runtimes-ref}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADP        & 391  &  130 &  222 &  339 &  486 \\
    OADP       & 15   &  136 &  230 &  350 &  499 \\
    ROADP      & 15   & 1211 & 1633 & 2023 & 2422 \\
    SP         & 16   &  130 &  222 &  339 &  486 \\
    RSP        & 15   & 1204 & 1624 & 2012 & 2409 \\
    AP         & 34   &  130 &  222 &  339 &  486 \\
    RAP        & 18   & 1204 & 1624 & 2012 & 2409 \\
    \hline
  \end{tabular}
  \caption{
    Study runtimes for simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-runtimes-study}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Study Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADP        & 0.00  & 0.00 & 0.00 & 0.00 & 0.00 \\
    OADP       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
    ROADP      & 0.23  & 0.15 & 0.12 & 0.08 & 0.08 \\
    SP         & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
    RSP        & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
    AP         & 0.46  & 0.37 & 0.31 & 0.26 & 0.22 \\
    RAP        & 3.02  & 2.93 & 2.54 & 2.15 & 1.91 \\
    \hline
  \end{tabular}
  \caption{
    Golden standard error for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-accuracy-gold}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADP        & 4.53  & 4.13 & 3.81 & 3.59 & 3.23 \\
    OADP       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
    ROADP      & 4.44  & 4.06 & 3.76 & 3.55 & 3.20 \\
    SP         & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
    RSP        & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
    AP         & 4.59  & 4.10 & 3.76 & 3.57 & 3.24 \\
    RAP        & 4.23  & 3.87 & 3.55 & 3.38 & 3.07 \\
    \hline
  \end{tabular}
  \caption{
    Geographic center error for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-accuracy-ctr}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADP        & 507  & 758 & 1011 & 1262 & 1549 \\
    OADP       &  73  & 108 &  156 &  189 &  282 \\
    ROADP      &  73  & 108 &  159 &  183 &  281 \\
    SP         &  44  &  64 &   84 &  103 &  131 \\
    RSP        &  43  &  63 &   83 &  101 &  129 \\
    AP         &  66  &  81 &  100 &  117 &  146 \\
    RAP        &  22  &  15 &   15 &   21 &   24 \\
    \hline
  \end{tabular}
  \caption{
    Study runtimes for the simulated data as the study size increases.
    The reference size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:mChg-runtimes-study}
\end{table}
